Young's Experiment

### 2020-11-12 fdtd/em.tex
WIP 12-Nov-2020.

plt.show()
\end{verbatim}
The results of the simulation are shown in figure \ref{f3}.
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-2}
\caption{Propagation of a Gaussian pulse in 2d}\label{f3}
\end{figure}
\bibliographystyle{plain}
\bibliography{em}
\end{document}

>>>>

plt.show()
\end{verbatim}
The results of the simulation are shown in figure \ref{f3}.
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-2}
\caption{Propagation of a Gaussian pulse in 2d}\label{f3}
\end{figure}

\section{Simulation of Young's experiment}\label{s9}
The code snippets in this section show how to simulate Young's experiment
using FDTD. This one is also a minor modification of the code in Houle in 
Sullivan's book \cite{sullivan2013electromagnetic}. We start with importing
all packages needed for the simulation. The first three are needed for the
numerical computations and generating the intensity plot. The remaining ones
are needed to store the plots as image files.
\begin{verbatim}
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.patches as patches 
import os
import time
\end{verbatim}

We define the physical constants used in the problem.
\begin{verbatim}
c = 3e8                 # The speed of light in vacuum.
kappa = 7e3             # The relative dielectric permittivity of the screen.
sigma = 0.1             # The conductivity of the screen.
epsilon_0 = 8.854e-12   # Permittivity of free space in SI units.
\end{verbatim}

The geometrical parameters of the problem are:
\begin{verbatim}
# The extent of the domain. **You can alter these parameters.**
xe = 100 # Width.
ye = 100 # Height.

# The central points in either directions.
yc = ye // 2 - 1
xc = xe // 2 - 1

# The location of the perfectly matching layer. 
# PML starts after these limits.
pml_dn = ye // 20           # starting vertical position after PML.
pml_up = ye - pml_dn - 1    # ending vertical position before PML.
pml_lf = xe // 20           # starting horizontal position after PML.
pml_rt = xe - pml_lf - 1    # ending horizontal position before PML.

# Slit geometry. **You can alter these parameters.**
slit_height = 2 # height of each slit (vertical measure).
slit_sep = 15 # separation between both slits (vertical measure).
screen_thinkness = 2 # thickness of the screen (horizontal measure).
\end{verbatim}

We now allocate space to store the two dimensional field vectors. We simulate
a TM wave propagating in vacuum until it hits the screen with two slits in 
it.
\begin{verbatim}
Ez = np.zeros((ye, xe))     # z-component of E field.
Dz = np.zeros((ye, xe))     # z-component of D field.
Hx = np.zeros((ye, xe))     # x-component of H field.
Hy = np.zeros((ye, xe))     # y-component of H field.
int_Ez = np.zeros((ye, xe)) # Time integral of E_z field.
int_Hx = np.zeros((ye, xe)) # Time integral of H_x field.
int_Hy = np.zeros((ye, xe)) # Time integral of H_y field.

Ez_in = np.zeros(xe) # z component of the incident E field.
Hx_in = np.zeros(xe) # x component of the incident H field.
\end{verbatim}

The simulation parameters and certain auxiliary variables are defined in the
code snippet below.
\begin{verbatim}
cell_size = 0.01 # Cell size in m
dt = cell_size / 2*c
# The dielectric profile.
dep_1 = np.ones((ye, xe))
dep_2 = np.zeros((ye, xe))
# The absorbing boundary conditions.
lob = [0, 0]
hib = [0, 0]

# **You can alter this parameter.**
t0 = 10 # Time for which the source is generated. 
# **You can alter this parameter.**
nsteps = 500  # Number of time steps.

# Vectors used for PML.
defy_2 = np.ones(ye)
defy_3 = np.ones(ye)
def2y_1 = np.zeros(ye)
def2y_2 = np.ones(ye)
def2y_3 = np.ones(ye)

defx_2 = np.ones(xe)
defx_3 = np.ones(xe)
def2x_1 = np.zeros(xe)
def2x_2 = np.ones(xe)
def2x_3 = np.ones(xe)

plotting_points = []
for n in range(1, nsteps+1):
    plotting_points.append({'label': 'a', 'num_steps': n, 'Ez_data': None}
\end{verbatim}

We now define the functions that do the FDTD computations. The first one
creates the dielectric profile of the domain. It initializes the portion of the
domain that has the screen with the screens $\sigma$ and $\kappa$.
\begin{verbatim}
def build_dielectric_profile():
  for j in range(pml_lf, pml_rt): # vertical range
      for i in range(pml_dn, pml_up): # horizontal range
          if point_on_screen(i, j): # If position is inside the dielectric                  
              dep_1[i, j] = 1 / (kappa + (sigma * dt / epsilon_0))
              dep_2[i, j] = (sigma * dt / epsilon_0)
\end{verbatim}

It uses another function that checks if a point is on the screen.
\begin{verbatim}
def point_on_screen(i, j):
    return (((i > 0 and i < yc - slit_height/2 - slit_sep/2) or 
              (i > yc - slit_sep/2 + slit_height/2 and 
               i < yc + slit_sep/2 - slit_height/2) or 
              (i > yc + slit_height/2 + slit_sep/2 and 
               i < ye)) 
              and (j > xc - screen_thinkness/2 and 
                   j < xc + screen_thinkness/2))
\end{verbatim}

The next function implements the absorbing boundary condition with a perfectly
matched layer.
\begin{verbatim}
def create_pml():
  for n in range(pml_dn + 1):    
      # auxp stands for auxiliary parameter, as defined in eqn (3.20) of 
      # Houle and Sullivan's book.
      auxp = ((pml_dn + 1 - n) / (pml_dn + 1)) ** 3
      auxp /= 3

      defy_2[n] = 1 / (1 + auxp)
      defy_2[ye - 1 - n] = 1 / (1 + auxp)
      defy_3[n] = (1 - auxp) / (1 + auxp)
      defy_3[ye - 1 - n] = (1 - auxp) / (1 + auxp)

      defx_2[n] = 1 / (1 + auxp)
      defx_2[xe - 1 - n] = 1 / (1 + auxp)
      defx_3[n] = (1 - auxp) / (1 + auxp)
      defx_3[xe - 1 - n] = (1 - auxp) / (1 + auxp)

      auxp = ((pml_dn + 0.5 - n) / (pml_dn + 1)) ** 3
      auxp /= 3

      def2y_1[n] = auxp
      def2y_1[ye - 2 - n] = auxp
      def2y_2[n] = 1 / (1 + auxp)
      def2y_2[ye - 2 - n] = 1 / (1 + auxp)
      def2y_3[n] = (1 - auxp) / (1 + auxp)
      def2y_3[ye - 2 - n] = (1 - auxp) / (1 + auxp)

      def2x_1[n] = auxp
      def2x_1[xe - 2 - n] = auxp
      def2x_2[n] = 1 / (1 + auxp)
      def2x_2[xe - 2 - n] = 1 / (1 + auxp)
      def2x_3[n] = (1 - auxp) / (1 + auxp)
      def2x_3[xe - 2 - n] = (1 - auxp) / (1 + auxp)
\end{verbatim}

The fields are computed in
\begin{verbatim}
def compute_fields(t): 
  if t % 100 == 0:
      print('|', end = '')
  elif t % 10 == 0:
      print('.', end = '')

  for j in range(1, xe):
      Ez_in[j] = Ez_in[j] + 0.5 * (Hx_in[j - 1] - Hx_in[j])

  # Absorbing Boundary Conditions
  Ez_in[0] = lob.pop(0)
  lob.append(Ez_in[1])  
  Ez_in[xe - 1] = hib.pop(0)
  hib.append(Ez_in[xe - 2])

  for j in range(1, xe):
      for i in range(1, ye):
          Dz[i, j] = defy_3[i] * defx_3[j] * Dz[i, j] + \
                      defy_2[i] * defx_2[j] * 0.5 * \
                      (Hy[i, j] - Hy[i - 1, j] -
                      Hx[i, j] + Hx[i, j - 1])

  # Source of electromagnetic waves.
  Ez_in[1] = (np.sin(2*np.pi*t/(2*t0)))**2

  for i in range(pml_dn, pml_up + 1):
      Dz[i, pml_lf] = Dz[i, pml_lf] + 0.5 * Hx_in[pml_lf - 1]
      Dz[i, pml_rt] = Dz[i, pml_rt] - 0.5 * Hx_in[pml_rt]

  for j in range(0, xe):
      for i in range(0, ye):
          Ez[i, j] = dep_1[i, j] * (Dz[i, j] - int_Ez[i, j])
          int_Ez[i, j] = int_Ez[i, j] + dep_2[i, j] * Ez[i, j]

  for j in range(0, xe - 1):
      Hx_in[j] = Hx_in[j] + 0.5 * (Ez_in[j] - Ez_in[j + 1])

  for j in range(0, xe - 1):
      for i in range(0, ye - 1):
          curl_e = Ez[i, j] - Ez[i, j + 1]
          int_Hx[i, j] = int_Hx[i, j] + curl_e
          Hx[i, j] = def2x_3[j] * Hx[i, j] + def2x_2[j] * \
                      (0.5 * curl_e + def2y_1[i] * int_Hx[i, j])

  for i in range(pml_dn, pml_up + 1):
      Hx[i, pml_lf - 1] = Hx[i, pml_lf - 1] + 0.5 * Ez_in[pml_lf]
      Hx[i, pml_rt] = Hx[i, pml_rt] - 0.5 * Ez_in[pml_rt]

  for j in range(0, xe):
      for i in range(0, ye - 1):
          curl_e = Ez[i, j] - Ez[i + 1, j]
          int_Hy[i, j] = int_Hy[i, j] + curl_e
          Hy[i, j] = def2y_3[i] * Hy[i, j] - def2y_2[i] * \
                      (0.5 * curl_e + def2x_1[j] * int_Hy[i, j])

  for j in range(pml_lf, pml_rt + 1):
      Hy[pml_dn - 1, j] = Hy[pml_dn - 1, j] - 0.5 * Ez_in[j]
      Hy[pml_up, j] = Hy[pml_up, j] + 0.5 * Ez_in[j]

  for pp in plotting_points:
      if t == pp['num_steps']:
          pp['Ez_data'] = np.copy(Ez)
\end{verbatim}

Although this code is enough to carry out the simulation, we need some
more functions to plot the results. The first one computes the average
intensity of the field at a fixed point.
\begin{verbatim}
def generate_avg_plot():
    left = int(ye * 0.2)
    right = int(ye * 0.8)
    x = [int(f) for f in np.linspace(left, right - 1, num = right - left)]

    t1 = int(nsteps * 0.8)
    t2 = nsteps - 1
    y0 = int(xe * 0.75) # Horizontal position where the filed is to be measured

    average = 0
    for i in range(t1, t2):
        average += abs(plotting_points[i]['Ez_data'][left:right, y0])

    plt.plot(x, average/(t2 - t1))
    plt.savefig('avg_intensity.png')
    plt.close()
\end{verbatim}

The next ones saves the field profiles at various time steps.
\begin{verbatim}
def create_rectangle(xy, width, height):
    clr = '#ffffff'
    return patches.Rectangle(xy, width, height, 
                             linewidth=1, edgecolor=clr, facecolor=clr)

def save_plots():       
    print('Saving images.')
    if not os.path.exists('figs'):
        os.makedirs('figs')

    pwd = os.getcwd()
    os.chdir('figs')

    xy1 = (xc - screen_thinkness/2, 0)
    xy2 = (xc - screen_thinkness/2, yc - slit_sep/2 + slit_height/2)
    xy3 = (xc - screen_thinkness/2, yc + slit_sep/2 + slit_height/2)

    for sp_no, pp in enumerate(plotting_points):  
        if sp_no % 100 == 0:
            print('|', end = '')
        elif sp_no % 10 == 0:
            print('.', end = '')

        r1 = create_rectangle(xy1, 
                              screen_thinkness, 
                              yc - slit_sep/2 - slit_height/2)
        r2 = create_rectangle(xy2, 
                              screen_thinkness, 
                              slit_sep - slit_height)
        r3 = create_rectangle(xy3, 
                              screen_thinkness, 
                              yc - slit_sep/2 - slit_height/2)

        ax = plt.figure().add_subplot(1, 1, 1)
        ax.add_patch(r1)
        ax.add_patch(r2)
        ax.add_patch(r3)

        plt.imshow(pp['Ez_data'], interpolation='bilinear', cmap='gist_heat')
        plt.savefig(f'figs_{str(sp_no + 1)}.png')
        plt.close()

    os.chdir(pwd)
\end{verbatim}
We call these functions in the main function.
\begin{verbatim}
def main():    
    build_dielectric_profile()
    create_pml()

    print('Computing the fields...')
    t1 = time.time()
    for t in range(1, nsteps + 1):
      compute_fields(t)

    t2 = time.time()
    print()
    print(f'Simulation took {round(t2 - t1, 2)} s.')
    generate_avg_plot()

    print('Saving the plots...')
    t1 = time.time()           
    save_plots()  
    t2 = time.time()
    print()
    print(f'Saving images took {round(t2 - t1, 2)} s.') 
\end{verbatim}

The intensity profile generated with $500$ steps of simulation are 
shown in figure \ref{f4}. Figure \ref{f5} shows the interference 
pattern.
\begin{figure}
\centering
\includegraphics[scale=0.5]{avg_intensity}
\caption{Intensity profile in Young's experiment}\label{f4}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.5]{figs_500}
\caption{Interference pattern in Young's experiment}\label{f5}
\end{figure}

\bibliographystyle{plain}
\bibliography{em}
\end{document}
